---
title: "CHDH SNA 5"
author: "[Johan Koskinen](https://psychologicalsciences.unimelb.edu.au/research/msps-research-groups/Social_Networks_Laboratory)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{css, echo=FALSE}
.question {
  background-color: lightpink;
  border: 3px solid red;
  font-weight: bold;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

-   Different types of data
-   Imperfect data

Load packages

```{r loadsna, results='hide', warning=FALSE, message=FALSE}
library("sna")
library("network")
library('RSiena')
```

# Two-mode data

Consider boards $i=1,\ldots,m$ and directors $j=1,\ldots,m$, and let
$$
\begin{equation*}  
X_{ij}= \left\{
\begin{array}{lr}
    1,&\text{if board } i \text{ has as a board member director } j\\
    0,&\text{else}
\end{array} 
\right. 
\end{equation*}
$$

## Data

A subset of the data on Swedish interlocking directors used in Koskinen and Edling (2012)

```{r}
SWE.b.t1 <- as.matrix(read.table("https://raw.githubusercontent.com/johankoskinen/CHDH-SNA/main/data/SWE-Borard-t1.txt"))

SWE.b.t2 <- as.matrix(read.table("https://raw.githubusercontent.com/johankoskinen/CHDH-SNA/main/data/SWE-Borard-t2.txt"))

```

Find out number of nodes in nodesets:

```{r}
nrsend <- dim( SWE.b.t1 )[1]
nrrec <- dim( SWE.b.t1 )[2]
```

To create bipartite network objects you need two node sets and must create the node sets too

```{r}
senders <- sienaNodeSet(nrsend, nodeSetName="senders")
receivers <- sienaNodeSet(nrrec, nodeSetName="receivers")
```

Define the dependent variable:

```{r}
myBipNet <- sienaDependent(array(c( SWE.b.t1,SWE.b.t2), dim=c(nrsend,nrrec,2)),# not square!
                      "bipartite",# this is the type (as opposed to one-mode)
                      nodeSet=c("senders","receivers")# associate rows and columns with nodesets
                      )
```

At this point we do not have any other variables:

```{r}
bipData <- sienaDataCreate(myBipNet,nodeSets=list(senders,receivers))
```

### What effects?

Get effects table for model specification:

```{r}
bipEffects <- getEffects(bipData)
```

### Descriptives

Generate initial descriptive output file:

```{r}
print01Report(bipData,modelname="bip1")
```

> We should really always look at the print-report

#### Four-cycles

```{r}
bipEffects <- includeEffects(bipEffects,cycle4,name="myBipNet")
```

#### Possible effects

```{r}
 bipEffects <- includeEffects(bipEffects,outActSqrt)
# Bypass bug related to initial values for bipartite network rates:
# bipEffects <- setEffect(bipEffects,Rate,name="leisure",initialValue=1.0,period=1,type="rate")
```

## Algorthm

```{r, results='hide',message=FALSE}
bip.algo <- sienaAlgorithmCreate( projname = 'bip.mod' )
```

Estimate!

```{r , results='hide',message=FALSE}
bip.mod.ans <- siena07( bip.algo,
                         data = bipData ,
                         effects =bipEffects ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

## Results

Check estimation results

```{r}
bip.mod.ans
```

## GOF

```{r}
(gofi0 <- sienaGOF(bip.mod.ans, IndegreeDistribution, verbose=TRUE, join=TRUE,
     varName="myBipNet"))
(gofo0 <- sienaGOF(bip.mod.ans, OutdegreeDistribution, verbose=TRUE, join=TRUE,
     varName="myBipNet"))
```

Note now that the indegree
$$
\sum_i x_{ij}
$$
will be, for each director $j$, the number of boards that choose director $j$

```{r}
plot(gofi0)
```


The outdegree
$$
\sum_j x_{ij}
$$
will be, for each board $i$, the number of directors the board has

```{r}
plot(gofo0)
```

> This simple model actually does relatively OK in terms of not creating boards that have too many or too few directors!

There is no simple way (to the best of my knowledge) of restricting the degree distribution. You may set the maximum using `MaxDegree=c(myBipNet=15)` in `sienaAlgorithmCreate`. There are some relevant effects as well.

# One-mode & two-mode

Consider now the Glasgow dataset and assume that we will model friendship $X$ alongside
$$
\begin{equation*}  
z_{ij}^{(r)}= \left\{
\begin{array}{lr}
    1,&\text{if pupil } i \text{ engages in category } j \text{ of activity } r\\
    0,&\text{else}
\end{array} 
\right. 
\end{equation*}
$$
for a number $r=1,\ldots,R$ different activities. The activities will be

- Leisure activities
- Music genres
- Different drugs


```{r}
temp <- tempfile()
download.file("http://www.stats.ox.ac.uk/~snijders/siena/Glasgow_bip_data.zip",temp)

# read the friendship data:
friendship1 <- as.matrix(read.table(unz(temp,"Glasgow-1-net.dat")))
friendship2 <- as.matrix(read.table(unz(temp,"Glasgow-2-net.dat")))

# now demographic info:
demographics <- as.matrix(read.table(unz(temp,"Glasgow-sex-birthyear.dat")))

# now the bipartite data:
leisure1 <- as.matrix(read.table(unz(temp,"GL-1-lsr.dat")))
leisure2 <- as.matrix(read.table(unz(temp,"GL-2-lsr.dat")))
music1 <- as.matrix(read.table(unz(temp,"GL-1-mus.dat")))
music2 <- as.matrix(read.table(unz(temp,"GL-2-mus.dat")))
drugs1 <- as.matrix(read.table(unz(temp,"GL-1-drg.dat")))
drugs2 <- as.matrix(read.table(unz(temp,"GL-2-drg.dat")))

unlink(temp)

```

### Recoding

Recode the behaviours into meaningful affiliation info:

```{r}
leisure1[leisure1 %in% c(2,3,4)] <- 0
# leisure1[leisure1 == 2] <- 1
leisure2[leisure2 %in% c(2,3,4)] <- 0
# leisure2[leisure2 == 2] <- 1
drugs1[drugs1 %in% c(1)] <- 0
drugs1[drugs1 %in% c(2,3,4)] <- 1
drugs2[drugs2 %in% c(1)] <- 0
drugs2[drugs2 %in% c(2,3,4)] <- 1
```

Recode valued friendship data & identify missings:

```{r}
friendship1[friendship1 == 2] <- 1
friendship1[friendship1 == 9] <- NA
friendship2[friendship2 == 2] <- 1
friendship2[friendship2 == 9] <- NA
```

We need to find out the number of nodes in different nodesets as well as the number of items in different scales:
```{r}
(nrpupils <- dim(leisure1)[1])
(nrleisureItems <- dim(leisure1)[2])
(nrmusicItems <- dim(music1)[2])
(nrdrugsItems <- dim(drugs1)[2])
```

## Node sets

We now have 4 different node sets. Define the different node sets for R:

```{r}
pupils <- sienaNodeSet(nrpupils, nodeSetName="pupils")
leisureItems <- sienaNodeSet(nrleisureItems, nodeSetName="leisureItems")
musicItems <- sienaNodeSet(nrmusicItems, nodeSetName="musicItems")
drugsItems <- sienaNodeSet(nrdrugsItems, nodeSetName="drugsItems")
```

## Dependent variables

We will format each affiliation network as a dependent variable

```{r}
leisure <- sienaDependent(array(c(leisure1, leisure2),
				dim=c(nrpupils, nrleisureItems,2)),
				"bipartite", nodeSet=c("pupils", "leisureItems"))
music <- sienaDependent(array(c(music1, music2),
				dim=c(nrpupils, nrmusicItems,2)),
				"bipartite", nodeSet=c("pupils", "musicItems"))
drugs <- sienaDependent(array(c(drugs1, drugs2),
				dim=c(nrpupils, nrdrugsItems,2)),
				"bipartite", nodeSet=c("pupils", "drugsItems"))
```

in addition to the dependent one-mode friendship network

```{r}
friendship <- sienaDependent(array(c(friendship1, friendship2),
				dim=c(nrpupils, nrpupils,2)), nodeSet="pupils")
```

## Covariate

Different node sets may have different covariates. Here we only use the sex of the pupils:

```{r}
sex.F <- coCovar(demographics[, 1], nodeSet="pupils")
```

## Data create

We have now defined all of the different variables and can combine them in an RSiena dataset

```{r}
bipData <- sienaDataCreate(friendship,# one-mode network
                           leisure, # pupil affiliation with leisure activity
                           music, # pupil affiliation with different music styles
                           drugs, # pupil affiliation with different drugs
                           sex.F, # sex (male/female; 0/1)
                           nodeSets=list(pupils, leisureItems, musicItems, drugsItems)# the different node sets
                           )

```

## Effects

Get effects table for model specification:
```{r}
bipEffects <- getEffects(bipData)
```

## Descriptives

Generate initial descriptive outputfile:
```{r}
print01Report(bipData, modelname="Bipartite-illustration")
```

Look at the generated output file "Bipartite-illustration.out" to see how RSiena interpreted the directives provided and to get an impression of the data (eyeball the degrees, look at occurrence of missings, look at Jaccard indices).

## Specify the model

When we specify the effects to include, pay special attention to the `name` variable

```{r}
bipEffects <- includeEffects(bipEffects, transTrip, name="friendship")
bipEffects <- includeEffects(bipEffects, from, name="friendship",
							interaction1="leisure")
bipEffects <- includeEffects(bipEffects, from, name="friendship",
							interaction1="music")
bipEffects <- includeEffects(bipEffects, from, name="friendship",
							interaction1="drugs")
bipEffects <- includeEffects(bipEffects, cycle4, name="leisure")
bipEffects <- includeEffects(bipEffects, cycle4, name="music")
bipEffects <- includeEffects(bipEffects, cycle4, name="drugs")
bipEffects <- includeEffects(bipEffects, to, name="leisure",
							 interaction1="friendship")
bipEffects <- includeEffects(bipEffects, to, name="music",
							 interaction1="friendship")
bipEffects <- includeEffects(bipEffects, to, name="drugs",
							 interaction1="friendship")
bipEffects
```

## Algorithm

Create algorithm object:
```{r}
bipModel <- sienaAlgorithmCreate(projname='bipartite-Glasgow-results')
```

**Estimate**!

```{r}
bipResults <- siena07(bipModel,
                      data=bipData,
                      effects=bipEffects,
                      batch = TRUE,
                      returnDeps = TRUE)
bipResults
```

## Results

Perhaps convergence is not satisfactory yet, so the model might be restimated in a further step:
```{r}
bipResults <- siena07(bipModel, data=bipData, effects=bipEffects,
							prevAns=bipResults,batch = TRUE, returnDeps=TRUE)
bipResults
```

# Multigroup analysis

A brief example of the data structure for running Bayesian hierarchical SAOMs. This requires the beta package `multiSiena` (used to be in `RSienaTest`). Best way is to download from
<https://www.stats.ox.ac.uk/~snijders/siena/multiSiena_1.2.20.tar.gz> (Mac) and <Download binary multiSiena_1.2.20.zip for installation under Windows> (PC), and then installing this from the Packages menue in RStudio.

```{r}
library(multiSiena)
```

## Data

We use a subset of a large dataset (Knecht, A., 2008. Friendship Selection and Friends' Influence Dynamics of Networks and Actor Attributes in Early Adolescence.')


```{r loaddata}
load(url("http://www.stats.ox.ac.uk/~snijders/siena/G21.RData"))
```

The main data are
```{r}
TwentyOne_Groups
class(TwentyOne_Groups)
length(TwentyOne_Groups)
```

This is a sienaGroup object, created by

```{r}
?sienaGroupCreate
```

The groups are the elements of `TwentyOne_Groups`. The first group is

```{r}
TwentyOne_Groups[[1]]
```

There is a group-level covariate `nn` that is `log(group size - 1)`

```{r}
group_n <- sapply(TwentyOne_Groups, function(x){length(x$nodeSets[[1]])})
group_n
```

## Print report

To further understand data, get the initial description

```{r}
print01Report(TwentyOne_Groups, modelname = 'Andrea_21Groups')
```

## Further material

Examples of how to specify a model for this multigroup object is provided at
<https://www.stats.ox.ac.uk/~snijders/siena/RscriptsienaBayes.r>
