---
title: "CHDH SNA 3"
author: "[Johan Koskinen](https://psychologicalsciences.unimelb.edu.au/research/msps-research-groups/Social_Networks_Laboratory)"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{css, echo=FALSE}
.question {
  background-color: lightpink;
  border: 3px solid red;
  font-weight: bold;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

-   All the effects!
-   Trouble shooting
-   Goodness of fit

Load packages

```{r loadsna, results='hide', warning=FALSE, message=FALSE}
library("sna")
library("network")
library('RSiena')
```

# All effects

## Data

Effects are data-dependent, so load data

```{r}
friend.data.w1 <- s501
friend.data.w2 <- s502
friend.data.w3 <- s503
drink <- s50a
smoke <- s50s
```

Define dependent and independent variables

```{r}
# format the dependent variables:
friendshipData <- array( c( friend.data.w1, 
                            friend.data.w2, 
                            friend.data.w3 ),
                         dim = c( 50, 50, 3 ) )
friendship <- sienaDependent(friendshipData)
drinking <- sienaDependent( drink, type = "behavior" )
smoking <- sienaDependent( smoke, type = "behavior" )
s50.big.data <- sienaDataCreate( friendship, smoking, drinking )
# chose effects
s50.big.effs <- getEffects( s50.big.data )
```

## Effects

The effects avaiable to us are in the object `s50.big.effs`. Have a look!

The most important columns are:

name | effectName | shortName | interaction1 | interaction2 | include | fix| test |          
:--------|:----------:| :----------:| :----------:| :----------:| :------:| :-----:| :----------:| 
Dependent variable that owns effect | description | what you use to define model | covariate 1 | covariate 2 | in model, T/F | T/F | T /F |

# Manual

The [manual]("https://www.stats.ox.ac.uk/~snijders/siena/RSiena_Manual.pdf") has a comprehensive list of possible effects

## Structural effects

Let us examine only the structural part of the previous model

```{r}
s50.big.effs <- includeEffects( s50.big.effs, transTrip, transRecTrip )
```

Set up estimation settings:

```{r, results='hide',message=FALSE}
s50.algo.struc.1<- sienaAlgorithmCreate( projname = 's50_struc.1' )
```

And estimate

```{r , results='hide',message=FALSE}
s50.est.struc.1 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.struc.1 )
```

What happens if we substitute `transTies` for `transRecTrip`

```{r}
s50.big.effs <- includeEffects( s50.big.effs,transRecTrip,include=FALSE )
s50.big.effs <- includeEffects( s50.big.effs, transTies )
```

And estimate

```{r , results='hide',message=FALSE}
s50.est.struc.2 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

```{r , results='hide',message=FALSE}
summary( s50.est.struc.2 )
```

Include all three
```{r}
s50.big.effs <- includeEffects( s50.big.effs,transRecTrip,include=TRUE )
```

And estimate

```{r , results='hide',message=FALSE}
s50.est.struc.3 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.struc.3 )
```

### GWESP

An alternative to having both `transTies` and `transTrip` is to use `GWESP`
```{r}
s50.big.effs <- includeEffects( s50.big.effs,transTies,transTrip,include=FALSE )
s50.big.effs <- includeEffects( s50.big.effs,gwespFF)
```
```{r , results='hide',message=FALSE}
s50.est.struc.4 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.struc.4 )
```

> What differences do we see between the different parametrizations?

## Selection effects

Looking at `s50.big.effs`, all effects with `name=friendship` and `interaction1=smoking` are social selection effects with respect to smoking.

```{r}
s50.effects.infl <- s50.big.effs
s50.effects.infl <- includeEffects( s50.effects.infl,# again, 'adding' to the effects
                                    egoX,# sender effect for alcohol
                                    egoSqX,# non-linear sender effect
                                    altX,# receiver effect for alcohol
                                    altSqX,# non-linear receiver effect
                                    diffSqX,# squared difference of alcohol ego and alcohol alter
                                    interaction1 = "drinking" # DV works the same as an IV on another DV
                                    )
```

Looking at `s50.big.effs`, all effects with `name=friendship` and `interaction1=smoking` are social selection effects with respect to drinking.

```{r}
s50.effects.infl <- includeEffects( s50.effects.infl,# we "add and effect" to our effects object
                               egoX,# the shortname here evokes that variable for 'ego' affects decision
                               interaction1 = "smoking" # the variable we want to interact the DV with
                               )
```

> Change or add a covariate effect on friendship of smoking

Estimate!
```{r , results='hide',message=FALSE}
s50.est.infl.1 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.effects.infl,
                         batch = TRUE,
                         returnDeps = TRUE )
```
	
Check results

```{r , results='hide',message=FALSE}
summary( s50.est.infl.1 )
```


## Influence effects

As before we specify

```{r}
s50.effects.infl.2 <- s50.big.effs
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    avAlt,# this is the shortname for "average alter"
                                    name="drinking",# name: what is the dependent variable
                                    interaction1 = "friendship" # the network is now "independent" variable
                                    )
```

But we can also add influence for `smoking` but let us try `totAlt` instead

```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    totAlt,# this is the shortname for "average alter"
                                    name="smoking",# name: what is the dependent variable
                                    interaction1 = "friendship" # the network is now "independent" variable
                                    )
```

The dependent variable can also be predictors of each other. Smoking as DV
```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    effFrom,# the effect of 'interaction1' on smoking
                                      name="smoking",# name: what is the dependent variable
                                    interaction1 = "drinking"# drinking is now "independent" variable
                                    )
```

Drinking as DV

```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    effFrom,# the effect of 'interaction1' on smoking
                                      name="drinking",# name: what is the dependent variable
                                    interaction1 = "smoking"# drinking is now "independent" variable
                                    )
```

### Interaction multiple variables

The effect of the total similarity of an egos alters drinking on the smoking of ego (smoking tot. sim. (friendship) x alter's drinking) requires two interactions
```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    totSimAltX,# the effect of 'interaction1' on smoking
                                    name="smoking",# name: what is the dependent variable
                                    interaction1 = "drinking",# drinking is now "independent" variable
                                    interaction2 = "friendship" 
                                    )
```


Estimate!
```{r , results='hide',message=FALSE}
s50.est.infl.2 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.effects.infl.2,
                         batch = TRUE,
                         returnDeps = TRUE )
```
	

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.infl.2 )
```


# Goodness-of-fit (GOF)
How do we know if the model is any good?

> You may simulate from the model to see if it replicates data

We already did this in `CHDH-SNA-2.Rmd`.

## Built in GOF routines

### Score-type test

What if we want to test if we should include `totSimAltX` for `drinking`?

The we add the effect but set `test=TRUE` and `fix=TRUE`

```{r}
s50.effects.infl.3 <- s50.effects.infl.2
s50.effects.infl.3 <- includeEffects( s50.effects.infl.3,# still augmenting effects structure
                                    totSimAltX,# the effect of 'interaction1' on smoking
                                    name="drinking",# name: what is the dependent variable
                                    interaction1 = "smoking",# drinking is now "independent" variable
                                    interaction2 = "friendship" ,
                                    test=TRUE, fix=TRUE
                                    )
```

The parameter is now fixed to 0

Estimate the model with`totSimAltX` for drinking set to 0

Estimate!
```{r , results='hide',message=FALSE}
(s50.est.infl.3 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.effects.infl.3,
                         batch = TRUE,
                         returnDeps = TRUE ))
```

We can now test
$$
H_0: \beta_{totSimAltX}=0
$$
against

$$
H_1: \beta_{totSimAltX}\neq0
$$

using a Score-type test:

```{r}
score.Test(s50.est.infl.3)
```

> This test only uses the simulated statistics for the effect and there is no need to actually estimate the parameter

# Fit of network measures

To test if the model replicates data well, we have to define what we want to look at.

The in- and out-degree distributions are already provided in the GOF routine

```{r}
(gofi0 <- sienaGOF(s50.est.infl.2, IndegreeDistribution, verbose=TRUE, join=TRUE,
     varName="friendship"))
(gofo0 <- sienaGOF(s50.est.infl.2, OutdegreeDistribution, verbose=TRUE, join=TRUE,
     levls=c(0:10,15,20),varName="friendship"))
(gof0.tc <- sienaGOF(s50.est.infl.2, TriadCensus, verbose=TRUE, join=TRUE,
     varName="friendship"))
```

## Inspecting GOF distributions

Plotting routines are defined for the GOF distributions. For example in-degree distribution

```{r}
plot( gofi0 )
```


# Trouble shooting

## Convergence issues

What if my model does not converge?

If converge t-statistics are close to $0.1$ in absolute value, restart using `prevAns`

Assume that we did not have enough iterations in Phase 3 **and** only 2 subphases in Phase 2
```{r, results='hide',message=FALSE}
s50.algo.struc.short<- sienaAlgorithmCreate( projname = 's50_struc.1', n3 =50, nsub=2)
```



Estimate!
```{r , results='hide',message=FALSE}
(s50.est.infl.3 <- siena07( s50.algo.struc.short,
                         data = s50.big.data ,
                         effects =s50.effects.infl.2,
                         batch = TRUE,
                         returnDeps = TRUE ))
```
