---
title: "CHDH SNA 4"
author: "[Johan Koskinen](https://psychologicalsciences.unimelb.edu.au/research/msps-research-groups/Social_Networks_Laboratory)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{css, echo=FALSE}
.question {
  background-color: lightpink;
  border: 3px solid red;
  font-weight: bold;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

-   Goodness of fit
-   Changing behaviour/social influence
-   All the effects!
-   Trouble shooting


Load packages

```{r loadsna, results='hide', warning=FALSE, message=FALSE}
library("sna")
library("network")
library('RSiena')
```

# Goodness-of-fit (GOF)
How do we know if the model is any good?

> You may simulate from the model to see if it replicates data

We already did this in `CHDH-SNA-3.Rmd` manually

## Built in GOF routines

If you estimate a model with `siena07` and set `returnDeps=TRUE`, the answer object returns the simulations from Phase 3. Let us try with the model for `s50` from `CHDH-SNA-3.Rmd`.

### Setting up s50

```{r loads50, results='hide'}
friend.data.w1 <- s501# s50 wave 1
friend.data.w2 <- s502# s50 wave 2
friend.data.w3 <- s503# s50 wave 3

smoke <- s50s# s50 smoking wave 1,2, and 3
smoke1.matrix <- as.numeric( smoke[,1]>1 )# dichotomise
smoke1 <- coCovar( smoke1.matrix )# format as constant covariate
friendshipData <- array( c( friend.data.w1, 
                            friend.data.w2, 
                            friend.data.w3 ),
                         dim = c( 50, 50, 3 ) )# paste together adjacency matrices in array
friendship <- sienaDependent(friendshipData)# designate network as dependent variable
s50data <- sienaDataCreate( friendship, smoke1) # create the siena data object
s50.effects <- getEffects(s50data) # create an effects structure
s50.effects <- includeEffects(s50.effects,# We "add an effect" to s50.effects
                              transTrip,# shortname of the effect
                              include=TRUE)
s50.effects <- includeEffects( s50.effects,# we "add an effect" to our effects object
                               egoX,# the shortname here evokes that variable for 'ego' affects decision
                               interaction1 = "smoke1" # the variable we want to interact the DV with
                               )
s50.effects <- includeEffects( s50.effects,
                               altX,# the shortname here evokes that variable for 'alter' affects decision of ego
                               interaction1 = "smoke1" # the variable we want to interact the DV with
                               )
s50.effects <- includeEffects( s50.effects,egoXaltX,interaction1 = "smoke1",include=FALSE)# this is homophily on smoking
s50algorithm.simple <- sienaAlgorithmCreate( projname = 's50_simple' )# We are only using defaults
```

Now estimate

```{r}
s50.hom.ans <- siena07( s50algorithm.simple,# estimation settings
                  data = s50data,# data object that knows DV and IV
                  effects = s50.effects,# the effects we specified
                  batch = TRUE,
                  returnDeps = TRUE )# IMPORTANT: we want to have access to networks simulated in Phase 3
```

## Fit of network measures

To test if the model replicates data well, we have to define what we want to look at.

The in- and out-degree distributions are already provided in the GOF routine

```{r}
(gofi0 <- sienaGOF(s50.hom.ans, IndegreeDistribution, verbose=TRUE, join=TRUE,
     varName="friendship"))
(gofo0 <- sienaGOF(s50.hom.ans, OutdegreeDistribution, verbose=TRUE, join=TRUE,
     levls=c(0:10,15,20),varName="friendship"))
(gof0.tc <- sienaGOF(s50.hom.ans, TriadCensus, verbose=TRUE, join=TRUE,
     varName="friendship"))
```

## Inspecting GOF distributions

Plotting routines are defined for the GOF distributions. For example in-degree distribution

```{r}
plot( gof0.tc )
```

# Dependent behaviour

We are now going to use drinking alcohol as **both** a changing covariate **and** a dependent variable.

## sienaDependent

Since the alcohol variable is a dependent variable, we use `sienaDependent` to format it (print to screen to get a quick view of the information stored)

```{r sienadepAlso, results='hide'}
drink <- s50a
drinking <- sienaDependent( drink, type = "behavior" )
```

## sienaDataCreate

We now have a dataset with two dependent variables so we need to create a new siena data object using `sienaDataCreate`:

```{r sienadatas50infl, results='hide'}
s50data.infl <- sienaDataCreate( friendship, smoke1, drinking )
```

## getEffects

With an additional variable, additional effects are available (**NB**:)

```{r}
s50.effects.infl <- getEffects(s50data.infl)
```


## As covariate

The dependent variable `drinking` can act as a covariate on the network. We are going to use the 'complicated', five-parameter, homophily parametrisation of Lomi and Snijder (2019):

```{r}
s50.effects.infl <- includeEffects( s50.effects.infl,# again, 'adding' to the effects
                                    egoX,# sender effect for alcohol
                                    egoSqX,# non-linear sender effect
                                    altX,# receiver effect for alcohol
                                    altSqX,# non-linear receiver effect
                                    diffSqX,# squared difference of alcohol ego and alcohol alter
                                    interaction1 = "drinking" # DV works the same as an IV on another DV
                                    )
```

## As a DV

The dependent variable `drinking` can act as a dependent variable for other variables, including the other dependent variables.

```{r}
s50.effects.infl <- includeEffects( s50.effects.infl,# still augmenting effects structure
                                    avAlt,# this is the shortname for "average alter"
                                    name="drinking",# name: what is the dependent variable
                                    interaction1 = "friendship" # the network is now "independent" variable
                                    )
```

## Other effects

We use the same social selection effects for smoking, the sender effect


```{r sendersmokeinfluence, results='hide'}
s50.effects.infl <- includeEffects( s50.effects.infl,# we "add and effect" to our effects object
                               egoX,# the shortname here evokes that variable for 'ego' affects decision
                               interaction1 = "smoke1" # the variable we want to interact the DV with
                               )
```

and the receiver effect


```{r receiversmokeSelec, results='hide'}
s50.effects.infl <- includeEffects( s50.effects.infl,
                               altX,# the shortname here evokes that variable for 'alter' affects decision of ego
                               interaction1 = "smoke1" # the variable we want to interact the DV with
                               )
```

and the homophily effect


```{r homophsmokeInfl, results='hide'}
s50.effects.infl  <- includeEffects( s50.effects.infl ,# we "add and effect" to our effects object
                               egoXaltX,# the shortname
                               interaction1 = "smoke1" # the variable we want to interact the DV with
                               )
```

## Structural effects

We will use *two* (2) different transitivity effects

```{r}
s50.effects.infl <- includeEffects( s50.effects.infl, transTrip, transRecTrip )
```

## Estimate

Set up estimation settings:

```{r s50influencealgo, results='hide',message=FALSE}
s50.algo.infl<- sienaAlgorithmCreate( projname = 's50_infl' )
```

And we can estimate the model (it will take a little longer)

```{r s50influenceest, results='hide',message=FALSE}
s50.est.infl <- siena07( s50.algo.infl,
                         data = s50data.infl,
                         effects = s50.effects.infl,
                         batch = TRUE,
                         returnDeps = TRUE )
```

## Results

We view the ANOVA table as before:

```{r}
summary(s50.est.infl)
```


Make a nicer listing of the results:
```{r}
siena.table(s50.est.infl, type="html", sig=TRUE)
```


# All effects

Let us explore more effects using the s50 data set.

## Data and model

Instead of using smoking as a constant covariate, we will define smoking as yet another dependent variable

```{r}
smoke <- s50s
```

Define dependent and independent variables

```{r}
# format the dependent variables:
friendshipData <- array( c( friend.data.w1, 
                            friend.data.w2, 
                            friend.data.w3 ),
                         dim = c( 50, 50, 3 ) )
friendship <- sienaDependent(friendshipData)
drinking <- sienaDependent( drink, type = "behavior" )
smoking <- sienaDependent( smoke, type = "behavior" )
s50.big.data <- sienaDataCreate( friendship, smoking, drinking )
# chose effects
s50.big.effs <- getEffects( s50.big.data )
```

## Effects

The effects avaiable to us are in the object `s50.big.effs`. Have a look!

The most important columns are:

name | effectName | shortName | interaction1 | interaction2 | include | fix| test |          
:--------|:----------:| :----------:| :----------:| :----------:| :------:| :-----:| :----------:| 
Dependent variable that owns effect | description | what you use to define model | covariate 1 | covariate 2 | in model, T/F | T/F | T /F |

## Manual

The [manual]("https://www.stats.ox.ac.uk/~snijders/siena/RSiena_Manual.pdf") has a comprehensive list of possible effects

## Structural effects

Let us examine only the structural part of the previous model

```{r}
s50.big.effs <- includeEffects( s50.big.effs, transTrip, transRecTrip )
```

Set up estimation settings:

```{r, results='hide',message=FALSE}
s50.algo.struc.1<- sienaAlgorithmCreate( projname = 's50_struc.1' )
```

And estimate

```{r , results='hide',message=FALSE}
s50.est.struc.1 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.struc.1 )
```

What happens if we substitute `transTies` for `transRecTrip`

```{r}
s50.big.effs <- includeEffects( s50.big.effs,transRecTrip,include=FALSE )
s50.big.effs <- includeEffects( s50.big.effs, transTies )
```

And estimate

```{r , results='hide',message=FALSE}
s50.est.struc.2 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

```{r , results='hide',message=FALSE}
summary( s50.est.struc.2 )
```

Include all three
```{r}
s50.big.effs <- includeEffects( s50.big.effs,transRecTrip,include=TRUE )
```

And estimate

```{r , results='hide',message=FALSE}
s50.est.struc.3 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.struc.3 )
```

### GWESP

An alternative to having both `transTies` and `transTrip` is to use `GWESP`
```{r}
s50.big.effs <- includeEffects( s50.big.effs,transTies,transTrip,include=FALSE )
s50.big.effs <- includeEffects( s50.big.effs,gwespFF)
```


```{r , results='hide',message=FALSE}
s50.est.struc.4 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.big.effs ,
                         batch = TRUE,
                         returnDeps = TRUE )
```

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.struc.4 )
```

> What differences do we see between the different parametrizations?

## Selection effects

Looking at `s50.big.effs`, all effects with `name=friendship` and `interaction1=smoking` are social selection effects with respect to smoking.

```{r}
s50.effects.infl <- s50.big.effs
s50.effects.infl <- includeEffects( s50.effects.infl,# again, 'adding' to the effects
                                    egoX,# sender effect for alcohol
                                    egoSqX,# non-linear sender effect
                                    altX,# receiver effect for alcohol
                                    altSqX,# non-linear receiver effect
                                    diffSqX,# squared difference of alcohol ego and alcohol alter
                                    interaction1 = "drinking" # DV works the same as an IV on another DV
                                    )
```

Looking at `s50.big.effs`, all effects with `name=friendship` and `interaction1=smoking` are social selection effects with respect to drinking.

```{r}
s50.effects.infl <- includeEffects( s50.effects.infl,# we "add and effect" to our effects object
                               egoX,# the shortname here evokes that variable for 'ego' affects decision
                               interaction1 = "smoking" # the variable we want to interact the DV with
                               )
```

> Change or add a covariate effect on friendship of smoking

Estimate!
```{r , results='hide',message=FALSE}
s50.est.infl.1 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.effects.infl,
                         batch = TRUE,
                         returnDeps = TRUE )
```
	
Check results

```{r , results='hide',message=FALSE}
summary( s50.est.infl.1 )
```


## Influence effects

As before we specify

```{r}
s50.effects.infl.2 <- s50.big.effs
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    avAlt,# this is the shortname for "average alter"
                                    name="drinking",# name: what is the dependent variable
                                    interaction1 = "friendship" # the network is now "independent" variable
                                    )
```

But we can also add influence for `smoking` but let us try `totAlt` instead

```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    totAlt,# this is the shortname for "average alter"
                                    name="smoking",# name: what is the dependent variable
                                    interaction1 = "friendship" # the network is now "independent" variable
                                    )
```

The dependent variable can also be predictors of each other. Smoking as DV
```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    effFrom,# the effect of 'interaction1' on smoking
                                      name="smoking",# name: what is the dependent variable
                                    interaction1 = "drinking"# drinking is now "independent" variable
                                    )
```

Drinking as DV

```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    effFrom,# the effect of 'interaction1' on smoking
                                      name="drinking",# name: what is the dependent variable
                                    interaction1 = "smoking"# drinking is now "independent" variable
                                    )
```

### Interaction multiple variables

The effect of the total similarity of an egos alters drinking on the smoking of ego (smoking tot. sim. (friendship) x alter's drinking) requires two interactions
```{r}
s50.effects.infl.2 <- includeEffects( s50.effects.infl.2,# still augmenting effects structure
                                    totSimAltX,# the effect of 'interaction1' on smoking
                                    name="smoking",# name: what is the dependent variable
                                    interaction1 = "drinking",# drinking is now "independent" variable
                                    interaction2 = "friendship" 
                                    )
```


Estimate!
```{r , results='hide',message=FALSE}
s50.est.infl.2 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.effects.infl.2,
                         batch = TRUE,
                         returnDeps = TRUE )
```
	

Check results

```{r , results='hide',message=FALSE}
summary( s50.est.infl.2 )
```


# More Goodness-of-fit (GOF)

## Built in GOF routines

### Score-type test

What if we want to test if we should include `totSimAltX` for `drinking`?

The we add the effect but set `test=TRUE` and `fix=TRUE`

```{r}
s50.effects.infl.3 <- s50.effects.infl.2
s50.effects.infl.3 <- includeEffects( s50.effects.infl.3,# still augmenting effects structure
                                    totSimAltX,# the effect of 'interaction1' on smoking
                                    name="drinking",# name: what is the dependent variable
                                    interaction1 = "smoking",# drinking is now "independent" variable
                                    interaction2 = "friendship" ,
                                    test=TRUE, fix=TRUE
                                    )
```

The parameter is now fixed to 0

Estimate the model with`totSimAltX` for drinking set to 0

Estimate!
```{r , results='hide',message=FALSE}
(s50.est.infl.3 <- siena07( s50.algo.struc.1,
                         data = s50.big.data ,
                         effects =s50.effects.infl.3,
                         batch = TRUE,
                         returnDeps = TRUE ))
```

We can now test
$$
H_0: \beta_{totSimAltX}=0
$$
against

$$
H_1: \beta_{totSimAltX}\neq0
$$

using a Score-type test:

```{r}
score.Test(s50.est.infl.3)
```

> This test only uses the simulated statistics for the effect and there is no need to actually estimate the parameter

# Trouble shooting

## Convergence issues

What if my model does not converge?

If converge t-statistics are close to $0.1$ in absolute value, restart using `prevAns`

Assume that we did not have enough iterations in Phase 3 **and** only 2 subphases in Phase 2
```{r, results='hide',message=FALSE}
s50.algo.struc.short<- sienaAlgorithmCreate( projname = 's50_struc.1', n3 =50, nsub=2)
```



Estimate!
```{r , results='hide',message=FALSE}
(s50.est.infl.3 <- siena07( s50.algo.struc.short,
                         data = s50.big.data ,
                         effects =s50.effects.infl.2,
                         batch = TRUE,
                         returnDeps = TRUE ))
```
